---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# mitoXpress

<!-- badges: start -->
<!-- badges: end -->

Vincent de Boer, March 23rd, 2023

🔬
🤖
🥼

This is an import and analysis script written for TRF mitoXpress experiments ran on a SpectraMax. Please have a good look at the data file and key file. 

## Install

For the script to run the following libraries are needed:

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(broom)
library(here)
library(janitor)
library(lubridate)
library(readxl)
library(bayestestR)
library(pspline)
library(colorspace)
library(ggrepel)
library(ggdist)

```

Also the renv lockfile in this repository is up to date with all packages that are used in this project.

This repository can be cloned to your own github page or the files can be copied to your own Rstudio project.

## Input

The data and key files are living in the folder `inst\extdata`.

### Data file

A data file starts its data column names on row three. The script skips the first two lines. The file contains a `Time` column and a `Temperature` column followed by one column for each `well`. The script removes rows that contain `NaN` and also removes the closing rows of the documents.

The data file is a tab-delimited text file. Apparently our Spectramax spits out txt files in UTF16 unicode format, which is somewhat outdated, since practically all unicode text files now are encoded as UTF8. This is likely the reason why Excel has difficulty reading the exported files from the spectramax properly.

The tab-delimited text column names are also renamed. Please notice that well names need to be three characters, so `A1` becomes `A01`. This conversion is also incorporated in the script since the plate reader outputs two and three digit well names. This is annoying when sorting data and plotting data.

The imported and processed data looks like this in R:

```{r load data, message=FALSE, warning=FALSE, include=FALSE}

filepath_mXp <- here::here("inst", "extdata", "evelien_test_expt6.txt")

mito_express_data <- filepath_mXp %>%
  readr::read_tsv(locale = readr::locale(encoding = "UTF-16LE"),
                  col_names = TRUE,
                  show_col_types = FALSE,
                  skip = 2) %>%
  filter(Time != "NaN",
         Time != "~End") %>%
  filter(!str_detect(Time, "Original")) %>%
  janitor::clean_names(case = "none") %>%
  rename(time = Time,
         temp = Temperature_C) %>%
  pivot_longer(c(-time, -temp),
               names_to = "well",
               values_to = "fluorescence") %>%
  mutate(well = well %>%
           map_chr(~paste0(
             str_extract(.x, "\\D"),
             str_extract_all(.x, "\\d") %>%
               pluck(1) %>%
               str_flatten() %>%
               str_pad(width = 2, pad = "0")))) %>%
  mutate(time = lubridate::hms(time),
         time_sec = lubridate::period_to_seconds(time)) %>%
  select(time, time_sec, temp, well, fluorescence) %>%
  drop_na()

```

```{r}
mito_express_data
```


### Key file

The `key file` is needed for the user to manually add meta info to the analysis. The most important is the choice for time frame for `slope` calculation. The time frame is given in seconds (`start` and `end`). Also `sample_type` and `group` information should be added. 

The key file is an excel file (.xlsx file).

The key file in R looks like:

```{r read key file, message=FALSE, warning=FALSE, include=FALSE}

filepath_keyfile <- here::here("inst", "extdata", "key_file_template.xlsx")

key_file <- readxl::read_excel(filepath_keyfile)
```

```{r}
key_file
```


## Plots

### raw data with smoothed spline

The first plot is the raw data traces with the smoothed spline.

```{r}
transformed <- readRDS(here::here("data", "transformed_4dpf_example.rds"))

transformed %>%
  ggplot(aes(x=time_sec, y = (fluorescence), color = well, group = group))+
  geom_point()+
  geom_line(aes(y = (zero_order), color = well, group = well),
            alpha = 0.5,
            linewidth = 2)+
  ggrepel::geom_label_repel(data = . %>% filter(time_sec == max(time_sec)),
                            aes(label = well),
                            show.legend = FALSE)+
  colorspace::scale_colour_discrete_divergingx(palette = "Geyser", rev = FALSE)+
  labs(y = "fluorescence (AU)",
       x = "time (sec)") +
  facet_wrap(~group, scales = "fixed")+
  theme_bw(base_size = 18)

```

### first order derivative of the smoothed spline

The second plot is the first derivative of the smoothed spline, it demosntrated were the maximum slope is.

```{r}
transformed %>%
  ggplot(aes(x=time_sec, y = first_order, color = well, group = group))+
  geom_point()+
  ggrepel::geom_label_repel(data = . %>% filter(time_sec == max(time_sec)),
                            aes(label = well),
                            show.legend = FALSE)+
  colorspace::scale_colour_discrete_divergingx(palette = "Geyser", rev = FALSE)+
  labs(y = "change in fluorescence (1st der.)",
       x = "time (sec)") +
  facet_wrap(~group, scales = "fixed")+
  theme_bw(base_size = 18)
```

### log transformed data

This is just a plot of the log transform of the smoothed spline.

```{r echo=FALSE}
transformed %>%
  filter(!group %in% c("empty", "cellfree neg.")) %>%
  ggplot(aes(x=time_sec, y = log(fluorescence), color = well, group = group))+
  geom_point()+
  geom_line(aes(y = log(zero_order), color = well, group = well),
            alpha = 0.5,
            linewidth = 2)+
  ggrepel::geom_label_repel(data = . %>% filter(time_sec == max(time_sec)),
                            aes(label = well),
                            show.legend = FALSE)+
  colorspace::scale_colour_discrete_divergingx(palette = "Geyser", rev = FALSE)+
  labs(y = "log fluorescence (AU)",
       x = "time (sec") +
  facet_wrap(~group, scales = "fixed")+
  theme_bw(base_size = 18)
```

### slopes 

Here the slopes are plotted for each well and plotted for each group as individual points and as point_interval (see also ggdist package). The dot is the median of the datapoints, the thick bar is where 95% of the data is and the thin bar is where 80% of the data is.

```{r}
transformed %>%
  filter(!group %in% c("empty")) %>%
  slice(1, .by = well) %>%
  ggplot(aes(x=group, y = log_slope_till_max, color = well, group = group))+
  geom_point(size = 4)+
  ggdist::stat_pointinterval(position = position_dodge(width = -0.4, preserve = "single"))+
  colorspace::scale_colour_discrete_divergingx(palette = "Geyser", rev = FALSE)+
  labs(y = "log slope (AU/1000sec)",
       x = "") +
  theme_bw(base_size = 18)+
  theme(panel.border = element_blank(),
        axis.line.y = element_line(),
        axis.line.x = element_line())
```

## Explanation of calculations

The data processing goed through several steps. It uses the `smooth.Pspline` from the `Psline` package. The smoothing parameters (spar = 0.8 and method = 2) are set. The zero-order and first_order splines are then processed and stored in the `transformed` tibble. Also, the area under the curve (AUC) is calculated and the original slope in the range from time_sec = 800 (s) to time_sec = 2800 (s) is calculated. These are not used anymore. 

To steps that are taken in the prcosssing are presented here:

```{r include=FALSE}
#location of digital plate image
filepath_tutorial_image <- here::here("mitoXpress_data_analysis_long.png")


```


`r sprintf("![filepath_tutorial_image](%s)", mitoXpress_data_analysis_long.png)`

`r filepath_tutorial_image`


